---
title: "county_stock"
format: html
editor: visual
---

C:/Users/ayip/Documents/GitHub/TEMPO/TEMPO.jl/data/passenger-projected-ldv-new-fuel-economy-ATB2024_1101-mid.csv

C:/Users/ayip/Documents/GitHub/TEMPO/TEMPO.jl/data/passenger-initial-ldv-fleet-fuel-economy-anl23-mid.csv

national stock

passenger-initial-ldv-stock-veh-vintage-moves-bev200.csv (default since decarb?)

county stock

|  |  |
|--------------------------------------|----------------------------------|
| passenger-initial-ALL-stock-veh_polk_2017_by_county_cleaned-aeovalid-bev200 | passenger-initial-ALL-stock-veh_polk_2017_by_county_cleaned-aeovalid |

So my hypothesis is that in these 4 counties (and possibly in too many counties), SUVs dominate the new sales, including fixed EV sales, in the near term, because of a mismatch in projected trends in ownership/demand vs. existing stock. this would pull up the energy use by weighting towards SUV. Lack of BEV200/400 might also be playing a role.

Whereas in national, it's likely more balanced/calibrated.

```{r}
library(tidyverse)
natlstock<-read_csv("C:/Users/ayip/OneDrive - NREL (1)/from eagle/oe/natl-household-stock.csv")
sjstock<-read_csv("C:/Users/ayip/OneDrive - NREL (1)/from eagle/oe/06085-household-stock.csv")
wystock<-read_csv("C:/Users/ayip/OneDrive - NREL (1)/from eagle/oe/56005-household-stock.csv")
oahunzstock<-read_csv("C:/Users/ayip/OneDrive - NREL (1)/from eagle/oe/oahu-nz-household-stock.csv")
oahumidstock<-read_csv("C:/Users/ayip/OneDrive - NREL (1)/from eagle/oe/oahu-mid-household-stock.csv")
tdmrunstock <- read_csv("c:/users/ayip/downloads/tdm-household-stock.csv")

allstock<-bind_rows(natlstock %>% mutate(Region = "national"), 
          sjstock %>% mutate(Region = as.character(Region)), 
          wystock %>% mutate(Region = as.character(Region)), 
          oahunzstock %>% mutate(Region = paste0(Region," NZ")), 
          oahumidstock %>% mutate(Region = paste0(Region, " Mid")),
          tdmrunstock %>% mutate(Region = paste0(Region, " tdm")))

# allstock %>%
# filter(Mode == "Personal_LDV",
#        #str_detect(Tech, "BEV|PHEV"),
#        Year < 2025 | Year %in% c(2025,2030,2040),
#        Vintage >= 2000
#        ) %T>%
#   skimr::skim() %>%
#   View()

```

```{r}
allstock %>%
  filter(Mode == "Personal_LDV",
       #str_detect(Tech, "BEV|PHEV"),
       Year < 2025 | Year %in% c(2025,2030,2040,2050),
       Vintage >= 2000
       ) %>%
#group_by(Region, Year, Mode, Class, Tech)
ggplot() +
  geom_histogram(aes(x=Vintage,weight=Vehicles,fill=Class), binwidth=1) +
  facet_grid(rows=vars(Region),cols=vars(Year), scales = "free_y") +
  #xlim(c(2000,2041)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "All-powertrain personal_ldv stock by vintage and class over simulation years")
ggsave("alltechstock_byvintageclass_oversimyears_forSJWYnatl.png", width = 10, height = 8)

allstock %>%
filter(Mode == "Personal_LDV",
       str_detect(Tech, "BEV|PHEV"),
       Year < 2025 | Year %in% c(2025,2030,2040,2050),
       Vintage >= 2000
       ) %>%
#group_by(Region, Year, Mode, Class, Tech)
ggplot() +
  geom_histogram(aes(x=Vintage,weight=Vehicles,fill=Class), binwidth=1) +
  facet_grid(rows=vars(Region),cols=vars(Year), scales = "free_y") +
  #xlim(c(2000,2041)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "BEV and PHEV personal_ldv stock by vintage and class over simulation years")
ggsave("EVstock_byvintageclass_oversimyears_forSJWYnatl.png", width = 10, height = 8)



allstock %>%
  mutate(new = case_when(Vintage > Year ~ "Vintage>Year", 
                         Vintage == Year ~ "Vintage=Year",
                         Vintage < Year ~ "Vintage<Year")) %>%
filter(Mode == "Personal_LDV",
       #str_detect(Tech, "BEV|PHEV"),
       Year < 2025 | Year %in% c(2025,2030,2040)
       ) %>%
#group_by(Region, Year, Mode, Class, Tech)
ggplot() +
  geom_col(aes(x=new,y=Vehicles,fill=Class)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(rows=vars(Region),cols=vars(Year), scales = "free_y") +
  labs(title = "All-powertrain personal_ldv stock by vintage bin and class over simulation years")
ggsave("alltechstock_byvintagebinclass_oversimyears_forSJWYnatl.png", width = 10, height = 8)

allstock %>%
  mutate(new = case_when(Vintage > Year ~ "Vintage>Year", 
                         Vintage == Year ~ "Vintage=Year",
                         Vintage < Year ~ "Vintage<Year")) %>%
filter(Mode == "Personal_LDV",
       #str_detect(Tech, "BEV|PHEV"),
       Year < 2025 | Year %in% c(2025,2030,2040,2050)
       ) %>%
#group_by(Region, Year, Mode, Class, Tech)
ggplot() +
  geom_bar(aes(x=new,weight=Vehicles,fill=Class), position = "fill") +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(rows=vars(Region),cols=vars(Year), scales = "free_y") +
  labs(title = "Class proportions of all-powertrain personal_ldv stock by vintage bin, over simulation years", y="Proportion of vehicles")
ggsave("alltechstock_byvintagebinclassproportion_oversimyears_forSJWYnatl.png", width = 10, height = 8)


allstock %>%
mutate(new = case_when(Vintage > Year ~ "Vintage>Year", 
                         Vintage == Year ~ "Vintage=Year",
                         Vintage < Year ~ "Vintage<Year")) %>%
  filter(Mode == "Personal_LDV",
       str_detect(Tech, "BEV|PHEV"),
       Year < 2025 | Year %in% c(2025,2030,2040,2050)
       ) %>%
#group_by(Region, Year, Mode, Class, Tech)
ggplot() +
  geom_col(aes(x=new,y=Vehicles,fill=Class)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(rows=vars(Region),cols=vars(Year), scales = "free_y") +
  labs(title = "BEV and PHEV personal_ldv stock by vintage bin and class over simulation years")
ggsave("EVstock_byvintagebinclass_oversimyears_forSJWYnatl.png", width = 10, height = 8)

```