---
title: "survival and uvm analysis"
author: "arthur yip"
format: html
editor: visual
---

# this is an older version

Request/ guidance from UTK

Define age as: (calendar year - model year). (MY are fixed. Lag age for comparison)

Let n(t,a,c) be the number of vehicles in operation of vehicle class c, of age a, in calendar year t.

The survival rate is n(t+1,a+1,c)/n(t,a,c) and the scrappage rate is 1-\[n(t+1,a+1,c)/n(t,a,c)\]

We would like as many ages as are available in the Experian data (in the IHS data this was 50 for 2020).

We need scrappage rates (or survival rates) by state and for the total U.S.

We grouped vehicles into three classes: 1. passenger cars, 2. SUVs + vans, 3. pickup trucks.

In our IHS data, we grouped the body types as follows: Passenger car = convertible, coupe, hatchback, sedan, wagon SUV + Van = SUV, Van Pickup = Pickup

We attempted to include only vehicles up to class 2a, i.e. =\< 8,500 lbs. GVW. However, it may not be possible to make precise GVW distinctions given the data available. Approximation is OK. Exclude 2b. Exclude weird stuff like limos and fire trucks.

Include I/O/fleetflag. split out later.

Ben: 80% of fleet vehicles are rental cars... other = business, govt, leasing...

??? is this true?

If breaking out by fuel type etc. then also provide combined and separate.... need "weights".

Data is from Eddie, processed via process_experian (Eagle), then process_experian_for_tempo.Rmd (stock section)

```{r}
library(tidyverse)

# processing and survival calculation functions
process_consecutive_years <- function(year1, year2) {
  data1 <- data_list[[as.character(year1)]] %>% mutate(CY = year1, CY_index = 1)
  data2 <- data_list[[as.character(year2)]] %>% mutate(CY = year2, CY_index = 2)
  
  combined_data <- bind_rows(data1, data2) %>%
    mutate(Age = CY - vehicle_model_year,
           Age1 = year1 - vehicle_model_year,
           Age2 = year2 - vehicle_model_year)
  return(combined_data)
}
calculate_survival <- function(pair_of_years, region = "natl", fleet_sep = F, tech_sep = F) {
  data <- processed_data_list[[as.character(pair_of_years)]]
  if (fleet_sep & region == "natl" & !tech_sep) {
    data <- data %>% 
      mutate(state_name = "USA", county_name = "USA", Tech = "ALL_TECHS")
  } else if (fleet_sep & region == "state" & !tech_sep) {
    data <- data %>% 
      mutate(county_name = "STATE", Tech = "ALL_TECHS")
  } else if (fleet_sep & region == "county" & !tech_sep) {
    data <- data %>% 
      mutate(Tech = "ALL_TECHS")
  } else if (!fleet_sep & region == "natl" & !tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS", state_name = "USA", county_name = "USA", Tech = "ALL_TECHS")
  } else if (!fleet_sep & region == "state" & !tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS", county_name = "STATE", Tech = "ALL_TECHS")
  } else if (!fleet_sep & region == "county" & !tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS", Tech = "ALL_TECHS")
  } else if (fleet_sep & region == "natl" & tech_sep) {
    data <- data %>% 
      mutate(state_name = "USA", county_name = "USA")
  } else if (fleet_sep & region == "state"& tech_sep) {
    data <- data %>% 
      mutate(county_name = "STATE")
  } else if (fleet_sep & region == "county"& tech_sep) {
    data <- data # all fields left alone!
  } else if (!fleet_sep & region == "natl"& tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS", state_name = "USA", county_name = "USA")
  } else if (!fleet_sep & region == "state"& tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS", county_name = "STATE")
  } else if (!fleet_sep & region == "county"& tech_sep) {
    data <- data %>% 
      mutate(regtype = "ALL_OWNERS")
  }
  
  data <- data %>%
    filter(!is.na(VehicleClass)) %>%
    group_by(regtype, Tech, state_name, county_name, Age1, VehicleClass, CY_index) %>%
    summarize(stock = sum(stock, na.rm=T)) %>%
    pivot_wider(names_from = "CY_index", values_from = "stock", names_prefix = "CY") %>%
    mutate(survival = CY2/CY1)
}


calculate_cumulative_survival <- function(survival) {
  survival %>%
    filter(Age1 >= 1) %>%
    group_by(regtype, Tech, state_name, county_name, VehicleClass) %>%
    arrange(regtype, Tech, state_name, county_name, VehicleClass, Age1) %>%
    mutate(cumulative_survival = cumprod(survival))
}

```

```{r}
# Directory where your CSV files are stored
directory <- "c:/users/ayip/OneDrive - NREL (1)/from eagle/"

# List to store data frames for each year
data_list <- list()

# Loop through years 2014 to 2022
for (year in 2014:2022) {
  # Construct the file path
  file_path <- paste0(directory, "exp_", year, "_stock_interim.csv")
  
  # Read the CSV file into a data frame
  data <- read_csv(file_path)
  
  # Store the data frame in the list
  data_list[[as.character(year)]] <- data
}

data_list <- data_list %>% map(~ .x %>% mutate(VehicleClass = case_when(yip_tempo_class %in% c("Compact","Midsize") ~ "Car",
                                  yip_tempo_class %in% c("SUV") ~ "SUV",
                                  yip_tempo_class %in% c("Pickup") ~ "Pickup"))) # i.e. excludes Class 2b

# Create list to store processed data for all pairs of consecutive years
processed_data_list <- list()

# Loop through consecutive pairs of years from 2014 to 2021
for (year in 2014:2021) {
  year_next <- year + 1
  # Process data for the current pair of years and store in the list
  processed_data_list[[paste0(year, year_next)]] <- process_consecutive_years(year, year_next)
}

year_pairs <- lapply(2014:2021, function(year) paste0(year, year + 1))

```

```{r}

national_survival_list_fleetsep <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "natl", fleet_sep = T, tech_sep = F)
cumulative_national_survival_list_fleetsep <- national_survival_list_fleetsep %>% map(calculate_cumulative_survival)

national_survival_list <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "natl", fleet_sep = F, tech_sep = F)
cumulative_national_survival_list <- national_survival_list %>% map(calculate_cumulative_survival)

state_survival_list_fleetsep <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "state", fleet_sep = T, tech_sep = F)
cumulative_state_survival_list_fleetsep <- state_survival_list_fleetsep %>% map(calculate_cumulative_survival)

state_survival_list <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "state", fleet_sep = F, tech_sep = F)
cumulative_state_survival_list <- state_survival_list %>% map(calculate_cumulative_survival)

county_survival_list_fleetsep <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "county", fleet_sep = T, tech_sep = F)
cumulative_county_survival_list_fleetsep <- county_survival_list_fleetsep %>% map(calculate_cumulative_survival)

county_survival_list <- year_pairs %>% setNames(year_pairs) %>% map(calculate_survival, region = "county", fleet_sep = F, tech_sep = F)
cumulative_county_survival_list <- county_survival_list %>% map(calculate_cumulative_survival)


# delivered to Greene, mid-June 2024
# national_survival_list %>% bind_rows(.id = "yearpair") %>% ungroup() %>% select(yearpair, Age1, VehicleClass, survival) %>% write_csv("c:/users/ayip/downloads/national_allyears_survival.csv")
# state_survival_list %>% bind_rows(.id = "yearpair") %>% ungroup() %>% select(yearpair, state_name, Age1, VehicleClass, survival) %>% write_csv("c:/users/ayip/downloads/state_allyears_survival.csv")

national_aggregate_allyears <- national_survival_list %>%
  bind_rows(.id = "yearpair") %>% 
  filter(yearpair != "20202021") %>%
  group_by(regtype, Age1, VehicleClass) %>% 
  summarize(CY2 = sum(CY2), CY1 = sum(CY1)) %>%
  mutate(survival = CY2/CY1) %>%
  filter(Age1 >= 1) %>%
  group_by(regtype, VehicleClass) %>%
  arrange(regtype, VehicleClass, Age1) %>%
  mutate(cumulative_survival = cumprod(survival))

national_aggregate_allyears_fleetsep <- national_survival_list_fleetsep %>%
  bind_rows(.id = "yearpair") %>% 
  filter(yearpair != "20202021") %>%
  group_by(regtype, Age1, VehicleClass) %>% 
  summarize(CY2 = sum(CY2), CY1 = sum(CY1)) %>%
  mutate(survival = CY2/CY1) %>%
  filter(Age1 >= 1) %>%
  group_by(regtype, VehicleClass) %>%
  arrange(regtype, VehicleClass, Age1) %>%
  mutate(cumulative_survival = cumprod(survival))




```

```{r plot-national-survival}

national_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
  filter(Age1 >= 0) %>% # & Age1 <= 50) %>%
  ggplot() +
  geom_line(aes(x=Age1, y=survival, group=VehicleClass, color=VehicleClass)) +
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  labs(y = "Annual Conditional Survival (ratio of vehicles in CY 2 of age X+1\nrelative to vehicles in CY 1 of age X)",
       title = "Empirical U.S. LDV conditional survival rates (year-over-year) for 2014-2022",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1")

#ggsave("c:/users/ayip/downloads/ann_surv_1422.png", width = 10, height = 8)

library(patchwork)
(cumulative_national_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
  
  geom_line(
    data = national_aggregate_allyears,
    mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    linetype = "dashed",
    linewidth = 1.5
    ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.2)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical U.S. LDV survival curves for 2014-2022",
       x = "Age in year 1")) /
  (cumulative_national_survival_list_fleetsep %>% 
  bind_rows(.id = "year_pair") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
    geom_line(
    data = national_aggregate_allyears_fleetsep,
    mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    linetype = "dashed",
    linewidth = 1.5
    ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.2)) +
  labs(y = "Cumulative survival rate",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1"))


#ggsave("c:/users/ayip/downloads/cum_surv_1422.png", width = 10, height = 8)

```

## NEXT TODO

- average out annual rates and get generalized state patterns over all year pairs

- plot state survivals by year pair 14-22

- 3 graphs for each of car/pu/suv

- geofacets

- 8 year-pair-specific curves on top of national avg and state avg

- first, a HI detour...

```{r plot-HI}
state_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
  filter(Age1 >= 0,
         state_name == "Hawaii") %>% # & Age1 <= 50) %>%
  ggplot() +
  geom_line(aes(x=Age1, y=survival, group=VehicleClass, color=VehicleClass)) +
  facet_wrap(~year_pair) + 
  labs(y = "Annual Conditional Survival (ratio of vehicles in CY 2 of age X+1\nrelative to vehicles in CY 1 of age X)",
       title = "Empirical Hawaiian LDV conditional survival rates (year-over-year) for 2014-2022",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1")

cumulative_state_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
  filter(state_name == "Hawaii") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
  
  geom_line(data = cumulative_national_survival_list %>% bind_rows(.id = "year_pair"),
            mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
            linetype = "dashed",
            linewidth = 1.5
            ) +
  
  facet_wrap(~year_pair) + 
  scale_y_continuous(limits = c(0,1.2)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical Hawaiian LDV survival curves for 2014-2022",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1")


(cumulative_state_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "Hawaii") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
    #   geom_line(
    # data = national_aggregate_allyears,
    # mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    # linetype = "dashed",
    # linewidth = 1.5
    # ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical Hawaii LDV survival curves for 2014-2022",
       x = "Age in year 1")) /
  (cumulative_state_survival_list_fleetsep %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "Hawaii") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
    #     geom_line(
    # data = national_aggregate_allyears_fleetsep,
    # mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    # linetype = "dashed",
    # linewidth = 1.5
    # ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1"))

ggsave("c:/users/ayip/downloads/")

# 2016-2017 fleet vehicles look weird in Hawaii
# 2015-2016 too - lots of hh imports... transfers ?

cumulative_state_survival_list_fleetsep[[3]] %>%
  filter(state_name == "Hawaii", VehicleClass == "Car")
```

```{r}
# check other state e.g. CA
(cumulative_state_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "California") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
      geom_line(
    data = national_aggregate_allyears,
    mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    linetype = "dashed",
    linewidth = 1.5
    ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical California LDV survival curves for 2014-2022",
       x = "Age in year 1")) /
  (cumulative_state_survival_list_fleetsep %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "California") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
        geom_line(
    data = national_aggregate_allyears_fleetsep,
    mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    linetype = "dashed",
    linewidth = 1.5
    ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1"))


```

# HI DATA

```{r}

HI_stateaggregate_allyears_anncondsurv <- state_survival_list %>%
  bind_rows(.id = "yearpair") %>% 
  #filter(yearpair != "20202021") %>%
  filter(yearpair != "20152016" & yearpair != "20162017") %>%
  group_by(regtype, Age1, VehicleClass) %>% 
  summarize(CY2 = sum(CY2), CY1 = sum(CY1)) %>%
  mutate(survival = CY2/CY1)

data_for_HI_stateagg_anncondsurv_for_tempo <- HI_stateaggregate_allyears_anncondsurv %>%
  filter(Age1 >= 0) %>%
  mutate(Annual_Survival = if_else(survival > 1, 1, survival)) %>% 
  ungroup() %>% 
  select(-regtype, -CY2, -CY1)

HI_stateagg_anncondsurv_for_tempo <- expand(tibble(), 
       Mode = "Personal_LDV",  
       Tech = c("ICEV_Gasoline", "HEV_Gasoline", "PHEV_25", "PHEV_50", "BEV_100", "BEV_300", "ICEV_NG", "FCEV"),
       Class = c("Compact","Midsize","Pickup","SUV"),
       Age = 0:99) %>%
  mutate(VehicleClass = if_else(Class %in% c("Compact","Midsize"), "Car", Class),
         Age1 = if_else(Age > 40, 40, Age)) %>%
  left_join(data_for_HI_stateagg_anncondsurv_for_tempo, by = c("Age1", "VehicleClass")) %>%
  select(-Age1, -VehicleClass, -survival)

survival_file_for_tempo <- read_csv("C:/Users/ayip/Documents/GitHub/TMM-dev/TEMPO.jl/data/survival/vehicle-survival-ldv_bytech-jb-class.csv") %>%
  filter(Mode != "Personal_LDV") %>%
  bind_rows(HI_stateagg_anncondsurv_for_tempo)

survival_file_for_tempo <- survival_file_for_tempo %>% mutate(Class = if_else(is.na(Class), "", Class),
                                                              Tech = if_else(is.na(Tech), "", Tech))

survival_file_for_tempo %>% filter(is.na(Mode) | is.na(Tech) | is.na(Class) | is.na(Age))

survival_file_for_tempo %>% write_csv("C:/Users/ayip/Documents/GitHub/TMM-dev/TEMPO.jl/data/survival/vehicle-survival-ldv_bytech-jb-class-LDV_HI.csv")

survival_file_for_tempo <- read_csv("C:/Users/ayip/Documents/GitHub/TMM-dev/TEMPO.jl/data/survival/vehicle-survival-ldv_bytech-jb-class-LDV_HI.csv")

survival_file_for_tempo %>% 
  filter(Mode == "Personal_LDV") %>%
  filter(Age >= 1) %>%
  group_by(Mode,Tech,Class) %>%
  arrange(Mode,Tech,Class,Age) %>%
  mutate(cumulative_survival = cumprod(Annual_Survival)) -> tempo_HI_surv

tempo_HI_surv %>%
  ggplot() +
  geom_line(aes(x=Age, y=cumulative_survival, group=interaction(Mode, Tech, Class), color =interaction(Mode, Tech, Class)))

```

```{r}

HI_aggregate_allyears <- HI_stateaggregate_allyears_anncondsurv %>%
  filter(Age1 >= 1) %>%
  group_by(regtype, VehicleClass) %>%
  arrange(regtype, VehicleClass, Age1) %>%
  mutate(cumulative_survival = cumprod(survival))



HI_aggregate_allyears_fleetsep <- state_survival_list_fleetsep %>%
  bind_rows(.id = "yearpair") %>% 
  filter(yearpair != "20162017") %>% # yearpair != "20202021" & 
  group_by(regtype, Age1, VehicleClass) %>% 
  summarize(CY2 = sum(CY2), CY1 = sum(CY1)) %>%
  mutate(survival = CY2/CY1) %>%
  filter(Age1 >= 1) %>%
  group_by(regtype, VehicleClass) %>%
  arrange(regtype, VehicleClass, Age1) %>%
  mutate(cumulative_survival = cumprod(survival))

# HI plots
# same as above
# thick and dashed are now HI avg, not natl avg?

(cumulative_state_survival_list %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "Hawaii") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
    #   geom_line(
    # data = HI_aggregate_allyears,
    # mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    # linetype = "dashed",
    # linewidth = 1.5
    # ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical Hawaii LDV survival curves for 2014-2022"
       )) /
  (cumulative_state_survival_list_fleetsep %>% 
  bind_rows(.id = "year_pair") %>%
    filter(state_name == "Hawaii") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
    #     geom_line(
    # data = HI_aggregate_allyears_fleetsep,
    # mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    # linetype = "dashed",
    # linewidth = 1.5
    # ) +
    
  facet_grid(cols = vars(year_pair), rows = vars(regtype)) + 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL",
       x = "Age in year 1"))

```

```{r HI-county}
HIcounty_allyears_fleetsep <- county_survival_list_fleetsep %>%
  bind_rows(.id = "yearpair") %>% 
  filter(yearpair != "20162017") %>% # yearpair != "20202021" & 
  filter(state_name == "Hawaii") %>%
  filter(county_name != "Kalawao" & county_name != "Kauai") %>%
  group_by(regtype, county_name, Age1, VehicleClass) %>% 
  summarize(CY2 = sum(CY2), CY1 = sum(CY1)) %>%
  mutate(survival = CY2/CY1)

HIcounty_aggregate_allyears_fleetsep <- HIcounty_allyears_fleetsep %>%
  filter(Age1 >= 1) %>%
  group_by(regtype, county_name, VehicleClass) %>%
  arrange(regtype, county_name, VehicleClass, Age1) %>%
  mutate(cumulative_survival = cumprod(survival))

HIcounty_aggregate_allyears_fleetsep %>% 
  bind_rows(.id = "year_pair") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass)) +
    
  geom_line(
    data = national_aggregate_allyears %>% ungroup() %>% select(-regtype),
    mapping = aes(x=Age1, y=cumulative_survival, group=VehicleClass, color=VehicleClass),
    linetype = "dashed",
    linewidth = 1.5
    #color = "black"
    ) +
      
  facet_grid(cols = vars(county_name, regtype), rows = vars(VehicleClass)) + # cols = vars(year_pair), 
  scale_y_continuous(limits = c(0,1.5)) +
  labs(y = "Cumulative survival rate",
       title = "Empirical Hawaii LDV survival curves (dashed = national average)",
       x = "Age in year 1")

```

```{r plot-state-survival}

library(geofacet)

state_survival_list %>%
  bind_rows(.id = "year_pair") %>%
  filter(Age1 >= 0) %>% # & Age1 <= 50) %>%
  filter(VehicleClass == "Car") %>%
  filter(year_pair == "20202021") %>%
  ggplot() +
  geom_line(aes(x=Age1, y=survival),
            col = "red") + #, group=VehicleClass, color=VehicleClass)) +
  labs(y = "Annual Conditional Survival (ratio of vehicles in CY 2 of age X+1\nrelative to vehicles in CY 1 of age X)") +
  facet_geo(facets = "state_name", grid = us_state_grid3, label = "name") +
  geom_line(data = national_survival_list[["20202021"]] %>% 
              filter(Age1 >= 0) %>% # & Age2018 <= 49) %>%
              filter(VehicleClass == "Car"),
            mapping = aes(x=Age1, y=survival),
            col = "black") #, group=VehicleClass, color=VehicleClass))


cumulative_state_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "Car") %>%
  ggplot() +
  facet_geo(facets = "state_name", grid = us_state_grid3, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "red") +
  labs(y = "Cumulative survival rates for cars by state",
       x = "Vehicle age",
       title = "Cumulative survival rates for cars by state (red) and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "Car"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black")

ggsave("c:/users/ayip/downloads/cum_surv_by_state_cars_1819.png", width = 10, height = 8)


cumulative_state_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "SUV") %>%
  ggplot() +
  facet_geo(facets = "state_name", grid = us_state_grid3, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "blue") +
  labs(y = "Cumulative survival rates for SUVs by state",
       x = "Vehicle age",
       title = "Cumulative survival rates for SUVs by state (blue) and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "SUV"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black")

ggsave("c:/users/ayip/downloads/cum_surv_by_state_SUVs_1819.png", width = 10, height = 8)


cumulative_state_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "Pickup") %>%
  ggplot() +
  facet_geo(facets = "state_name", grid = us_state_grid3, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "orange") +
  labs(y = "Cumulative survival rates for light-duty pickup trucks by state",
       x = "Vehicle age",
       title = "Cumulative survival rates for light-duty pickup trucks by state (orange) and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "Pickup"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black")

ggsave("c:/users/ayip/downloads/cum_surv_by_state_pus_1819.png", width = 10, height = 8)




```

```{r fig.width = 6, fig.height = 12}
library(geofacet)

county_survival <- exp1819 %>%
  filter(!is.na(VehicleClass)) %>%
  group_by(FIPS_5, county_name, state_name, Age2018, VehicleClass, CY) %>% # lump all else e.g. fuel type, I/O/F
  summarize(stock = sum(stock, na.rm=T)) %>%
  pivot_wider(names_from = "CY", values_from = "stock", names_prefix = "CY") %>%
  mutate(survival = CY2019/CY2018)
  
cumulative_county_survival <- county_survival %>% 
  filter(Age2018 >= 1) %>%
  group_by(FIPS_5, county_name, state_name, VehicleClass) %>%
  arrange(FIPS_5, VehicleClass, Age2018) %>%
  mutate(cumulative_survival = cumprod(survival))

cumulative_county_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "Car") %>%
  filter(state_name == "California") %>%
  filter(cumulative_survival <= 2) %>%
  ggplot() +
  facet_geo(facets = "county_name", grid = us_ca_counties_grid1, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "red") +
  labs(y = "Cumulative survival rates for cars by county in California",
       x = "Vehicle age",
       title = "Cumulative survival rates for cars by county (red), state (purple), and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "Car"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black") +
    geom_line(data = cumulative_state_survival %>% 
              filter(VehicleClass == "Car") %>%
                filter(state_name == "California"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "purple")

ggsave("c:/users/ayip/downloads/cum_surv_by_county_CA_cars_1819.png", width = 8, height = 12)


cumulative_county_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "SUV") %>%
  filter(state_name == "California") %>%
  filter(cumulative_survival <= 2) %>%
  ggplot() +
  facet_geo(facets = "county_name", grid = us_ca_counties_grid1, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "blue") +
  labs(y = "Cumulative survival rates for SUVs by county in California",
       x = "Vehicle age",
       title = "Cumulative survival rates for SUVs by county (blue), state (purple), and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "SUV"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black") +
      geom_line(data = cumulative_state_survival %>% 
              filter(VehicleClass == "SUV") %>%
                filter(state_name == "California"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "purple")

ggsave("c:/users/ayip/downloads/cum_surv_by_county_CA_SUVs_1819.png", width = 8, height = 12)


cumulative_county_survival %>% 
  ungroup() %>%
  filter(VehicleClass == "Pickup") %>%
  filter(state_name == "California") %>%
  filter(cumulative_survival <= 2) %>%
  ggplot() +
  facet_geo(facets = "county_name", grid = us_ca_counties_grid1, label = "name") +
  geom_line(aes(x=Age2018, y=cumulative_survival), 
            col = "orange") +
  labs(y = "Cumulative survival rates for light-duty pickup trucks by county in California",
       x = "Vehicle age",
       title = "Cumulative survival rates for light-duty pickup trucks by county (orange), state (purple), and national (black)",
       subtitle = "based on observed differences in CY2019 and CY2018 vehicle stocks",
       caption = "Source: Derived from Experian registration data, via Arthur Yip, NREL") +
  geom_line(data = cumulative_national_survival %>% 
              filter(VehicleClass == "Pickup"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "black") +
        geom_line(data = cumulative_state_survival %>% 
              filter(VehicleClass == "Pickup") %>%
                filter(state_name == "California"),
            mapping = aes(x=Age2018, y=cumulative_survival),
            col = "purple")

ggsave("c:/users/ayip/downloads/cum_surv_by_county_CA_pus_1819.png", width = 8, height = 12)



```

```{r}

greene_stock_data_request <- data_list$`2022` %>%
  mutate(Tech = if_else(tempo_fuel_type_with_range == "HEV", "HEV", Tech)) %>%
  group_by(state_name, Tech, vehicle_model_year, VehicleClass) %>%
  summarize(stock = sum(stock))
greene_stock_data_request
greene_stock_data_request %>% write_csv("c:/users/ayip/downloads/state_2022_stocks_with_HEV.csv")

```

# state / national post-analysis

```{r}
greene <- read_excel("c:/users/ayip/OneDrive - NREL (1)/Attachments/StateNationalSurvivalRateDifferences12Dec25NewMethod.xlsx", sheet = "NetImports")
greene

vehstockdata_forweightinggreene <- read_csv("c:/users/ayip/downloads/prototype.csv") %>%
  mutate(VehType = case_when(class_tempo == "SUV" ~ "SUV",
                             class_tempo == "Pickup" ~ "Pickup",
                             TRUE ~ "Car")) %>%
  filter(age <= 8 & age >= 2) %>% # same population cohort that Greene analyzed
  left_join(summary_stats_x, by = "Region") %>%
  group_by(STNAME, VehType) %>%
  summarize(stock = sum(vehicle_count), .groups = "drop_last") 

greene_intermediate <- greene %>% 
  select(VehType = VehType...1, STNAME = State...2, NetImpExpMean1421 = Mean, 
         NetImpExp19 = `2019...9`, 
         NetImpExp18 = `2018...8`,
         NetImpExp17 = `2017...7`,
         NetImpExp16 = `2016...6`,
         NetImpExp15 = `2015...5`) %>%
  filter(!is.na(VehType)) %>%
  left_join(vehstockdata_forweightinggreene, by = c("VehType", "STNAME")) %>%
  mutate(NetImpExpMean1519_stock = (NetImpExp19+NetImpExp18+NetImpExp17+NetImpExp16+NetImpExp15)/5*stock*0.92, # deflating stock for older year data
         NetImpExpMean1421_stock = NetImpExpMean1421 * stock * 0.93, # deflating stock for older year data
         NetImpExp19_stock = NetImpExp19 * stock * 0.95,
         NetImpExp18_stock = NetImpExp18 * stock * 0.94,
         NetImpExp17_stock = NetImpExp17 * stock * 0.93,
         NetImpExp16_stock = NetImpExp16 * stock * 0.92,
         NetImpExp15_stock = NetImpExp15 * stock * 0.91)

greene_intermediate

greene_intermediate %>%
  group_by(VehType) %>%
  summarize(national_net_mean1421 = sum(NetImpExpMean1421_stock, na.rm=T),
            national_net_mean1519 = sum(NetImpExpMean1519_stock, na.rm=T),
            national_net_19 = sum(NetImpExp19_stock, na.rm=T),
            national_net_18 = sum(NetImpExp18_stock, na.rm=T),
            national_net_17 = sum(NetImpExp17_stock, na.rm=T),
            national_net_16 = sum(NetImpExp16_stock, na.rm=T),
            national_net_15 = sum(NetImpExp15_stock, na.rm=T),
            national_stock = sum(stock * 0.93, na.rm=T)) %>%
  mutate(pct_net = national_net_mean1519/national_stock)

```

```{r}



library(tidyverse)
library(leaflet)
library(sf)
library(tigris) # Provides easy access to US Census shapefiles
library(RColorBrewer) # Provides color palettes
options(tigris_use_cache = TRUE) # Cache shapefiles for future use
#library(mapview)

# 1. Load data
#greene_intermediate

for (VehTypeCurrent in c("Pickup", "SUV", "Car")) {

# 2. Get US states spatial data
# Use the tigris package to get state boundaries, including shifted AK/HI insets
states_sf <- states(cb = TRUE, resolution = "20m") %>%
  filter(!STUSPS %in% c("PR", "VI", "GU", "AS", "MP")) %>%
  rename(STNAME = NAME) %>% 
  left_join(greene_intermediate %>% filter(VehType == VehTypeCurrent), by = "STNAME") %>% 
  arrange(STNAME)

# contiguous <- states_sf %>% filter(!STNAME %in% c("Alaska", "Hawaii"))
# alaska <- states_sf %>% filter(STNAME == "Alaska")
# hawaii <- states_sf %>% filter(STNAME == "Hawaii")
# 
# # Transform Alaska
# alaska_transformed <- alaska %>%
#   st_transform(3338) %>%  # Alaska Albers
#   st_geometry() %>%
#   st_cast("MULTIPOLYGON") %>%
#   `*`(0.35) %>%  # Scale to 35%
#   `+`(c(-1500000, -4300000)) %>%  # Move position
#   st_set_crs(3338) %>%
#   st_transform(st_crs(states_sf))
# 
# alaska$geometry <- alaska_transformed
# 
# # Transform Hawaii  
# hawaii_transformed <- hawaii %>%
#   st_transform(4135) %>%  # Hawaii projection
#   st_geometry() %>%
#   st_cast("MULTIPOLYGON") %>%
#   `*`(1.5) %>%  # Scale to 150%
#   `+`(c(5400000, -1400000)) %>%  # Move position
#   st_set_crs(4135) %>%
#   st_transform(st_crs(states_sf))
# 
# hawaii$geometry <- hawaii_transformed
# 
# # Combine back together
# states_sf_transformed <- bind_rows(contiguous, alaska, hawaii)
states_sf_transformed <- states_sf

# 3. Define color palette and popups
pal <- colorNumeric(
  palette = "RdBu",
  #domain = states_sf$NetImpExpMean1421_stock
  domain = c(-150000, 150000)
  )

# Prepare labels for popups (using HTML for formatting)
state_labels <- paste(
  "<strong>", states_sf_transformed$STNAME, "</strong><br/>",
  "Net Vehicle (Age 2-8) Import/Export: ", round(states_sf_transformed$NetImpExpMean1519_stock, 0)
) %>% lapply(htmltools::HTML)


# 4. Create the Leaflet map
m <- leaflet(states_sf_transformed) %>%
    setView(lng = -96, lat = 37.8, zoom = 4) %>%  # Center on USA
  setMaxBounds(lng1 = -125, lat1 = 24, lng2 = -66, lat2 = 50) %>%
  #addTiles() %>% # Add default OpenStreetMap map tiles
  addPolygons(
    fillColor = ~pal(NetImpExpMean1519_stock),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE),
    label = state_labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLabelOnlyMarkers(
    data = states_sf_transformed,
    lng = ~st_coordinates(st_centroid(geometry))[,1],
    lat = ~st_coordinates(st_centroid(geometry))[,2],
    label = ~round(states_sf_transformed$NetImpExpMean1519_stock, 0), 
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "center",
      textOnly = TRUE,
      style = list(
        "color" = "black",
        "font-weight" = "bold",
        "font-size" = "12px"
      )
    )
    ) %>%
      addLegend(pal = pal, values = ~NetImpExpMean1519_stock, opacity = 0.7, title = paste("Net",VehTypeCurrent,"(Age 2-8) Import/Export"),
            position = "bottomright") # Add legend

# Display the map
m

htmlwidgets::saveWidget(m, file = paste0("c:/users/ayip/downloads/",VehTypeCurrent,".html"), selfcontained = TRUE)
#mapview::mapshot(m, file = paste0("c:/users/ayip/downloads/",VehTypeCurrent,".png"))
}
```